---
title: "ELISA logistic regression analysis"
output:
  html_document:
    df_print: paged
---

Use this script to run logistic regression on absorbance values

```{r}
options(warn=-1)
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

```{r}
# library(ELISAtools)
library(drc)
library(tidyverse)
library(readxl)
library(ggpubr)

# could just make this a project...
data_sheet <- file.path("data", "2022-04-13 SFLT.xlsx") # 2022-03-11 sflt elisa.xlsx
output_dir <- file.path("output", 
                        str_extract(string = basename(data_sheet), 
                                    pattern = "[0-9]+-[0-9]+-[0-9]+"))
dir.create(output_dir, recursive = T, showWarnings = F)


standard_dat <- read_excel(data_sheet, sheet = 1)  %>%
  # assuming that Concentration of STandard is the 1st column
  mutate(log2concentration = log2(.[[1]]),
         Absorbance = .[[2]])
results <- read_excel(data_sheet, sheet = 2) %>%
  mutate(Samples = .[[1]],
         Absorbance = .[[2]])

# logistic regression model; estimate slope, ED50, and upper and lower bounds using non-linear method
# logic behind this model: where we model the ED50 in log space,  four-parameter log-logistic model may be preferred for dose-response analysis involving very small datasets (<15–20) where a normally distributed parameter estimate may more reasonably be assumed on a logarithm-transformed dose scale than the original dose scale.

# drc paper: https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0146021#sec001

fit <- drm(formula = Absorbance ~ log2concentration, 
           data = standard_dat, 
           fct=LL.4(names=c("Slope", "Lower", "Upper", "ED50")),
           logDose = 2 # log 2 doses
)

# print the fit summary
print(summary(fit))
```

```{r}
# this is teh adjusted absorbances of each of the samples; we want to predict their concentration based off the standard curve
response <- results$Absorbance

# in the order of the samples
DOSEx_temp <- ED(fit,response,type="absolute",display=F) %>% as_tibble()

DOSEx <- DOSEx_temp %>% as.data.frame() %>%
  mutate(up = Estimate + `Std. Error`,
         down = Estimate - `Std. Error`) %>%
  mutate(across(.cols = everything(), log2))



combined_res <- bind_cols(results, DOSEx_temp, 
                          DOSEx %>% rename_with(.fn = ~ str_c(.x, "_log2"), .cols = everything())) %>%
  dplyr::rename(`fit_to_curve_concentration_estimate` = Estimate,
                `log2_fit_to_curve_concentration_estimate` = Estimate_log2)

print(combined_res %>% select(1:3))
```

# RESULTS
## Note: Estimated concentrations are the same unit as the standard's concentrations
```{r plot}
print(DOSEx)
plot_fn <- file.path(output_dir, "results_plot.pdf")

# print the plot in the markdown document
plot(fit, 
     xlim = c(min(DOSEx[,1], na.rm = T), max(DOSEx[,1], na.rm = T)),
     ylim = c(min(response, na.rm = T), max(response, na.rm = T)), main = "Plot of absorbance x concentration\nFitted sample values overlayed") +
  points(y=response,x=DOSEx[,1],col="blue",pch=19,cex=0.75) +
  # With error bars
  arrows(DOSEx[,1], response, DOSEx$up, response, length=0.05, angle=90, lwd=1,col="blue") +
  arrows(DOSEx[,1], response, DOSEx$down, response, length=0.05, angle=90, lwd=1,col="blue")


# save the plot
save_plot <- function() {
  pdf(file = plot_fn, width = 6, height = 6)
  plot(fit, 
       xlim = c(min(DOSEx[,1], na.rm = T), max(DOSEx[,1], na.rm = T)),
       ylim = c(min(response, na.rm = T), max(response, na.rm = T)), main = "Plot of absorbance x concentration\nFitted sample values overlayed") +
    points(y=response,x=DOSEx[,1],col="blue",pch=19,cex=0.75) +
    # With error bars
    arrows(DOSEx[,1], response, DOSEx$up, response, length=0.05, angle=90, lwd=1,col="blue") +
    arrows(DOSEx[,1], response, DOSEx$down, response, length=0.05, angle=90, lwd=1,col="blue")
  dev.off()
}
save_plot()
```
(`r knitr::fig_chunk('plot', 'png')`)


# Results written to this directory:
```{r}
message("writing to output directory:")
message(output_dir)
write_tsv(combined_res %>% select(Samples, Absorbance, fit_to_curve_concentration_estimate), 
          file = file.path(output_dir, "results.tsv"))
options(warn=0)
```


# More about *drc* package and 4-param logit Dose Response model

The full specification of a statistical dose-response (regression) model involves specifying how the mean is described by a parametric function of dose as well as specifying assumptions about the distribution of the response.

We parameterize this model using a unified structure with a coefficient $b$, denoting the **steepness** of the dose-response curve, $c$ and $d$, the lower and upper asymptotes or limits of the response, and $\tilde{e}$ is the $log_2$ of the effective dose (ED50). Note: These are the $\beta$'s

$$ f(x, (b, c, d, \tilde{e})) = c + \frac{d - c}{1 + exp(b(log(x) - \tilde{e}))} $$
This 4-param log-logistic regression model is preferred with very small datasets (<15–20) because without the logarithm, $e$ or the ED50 may not be normally distributed at  the original dose scale.


## More information about the statistical model being used
Let $y$ denote an observed response value corresponding to a dose value $x$ greater than or equal to 0. The values of $y$ are often positive but may take arbitrary positive or negative values (for instance, could be log2 transformed, as in our case). Furthermore, we will assume that observation of $y$ is subject to sampling variation, necessiating the specification of a statistical model describing the random variation. Specifically, we will focus on characterizing the mean of $y$ (denoted $E(y)$ below) in terms of a model function $f$ that depends on the dose $x$:

$$E(y) = f(x,\beta)$$

So, for a given dose $x$, the corresponding observed response values will be distributed around $f(x, \beta)$. The function $f$ is completely known as it reflects the assumed relationship between $x$ and $y$, except for the values of the model parameters $\beta = (\beta1, …, \beta p)$, which will be estimated from the data to obtain the best fitting function (**these are the 4 parameters being estimated for our model above**, $b$, $c$, $d$, and $\tilde{e}$). The remaining distributional assumptions on $y$ will depend on the type of response (which is continous for us). 

